---
title: "National Weather Service Storm Data Dangers and Costs"
author: "Justin Meyer"
date: "February 5, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis of Analysis

This analysis project uses National Weather Service storm data to compare injuries/fatalities and costs associated with storm damage by the type of storm that occurred (thunderstorm, flood, tornado, fire, etc.). The data are available from 1950 to 2011 though the data are less complete in earlier years. This analysis shows that tornadoes are the most dangerous type of storm and tornado/thunderstorm wind/hail are the most costly.

## Data Processing

The categorization of the type of storm is critical to this analysis but unfortunately the values that are used to classify the storms are inconsistent, for example, with "TSTM WIND" and "THUNDERSTORM WIND" both indicating a thunderstorm wind. In order to be able to classify the storm type accurately I merged some types as shown in the code below.

```{r}

# Get data
storms <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")

# Considering only storm types with more than 500 records, recode storm types that appear to be the same into the most common type.

storms$EVTYPE[storms$EVTYPE == "THUNDERSTORM WINDS" |
                      storms$EVTYPE == "TSTM WIND"] <- "THUNDERSTORM WIND"

storms$EVTYPE[storms$EVTYPE == "STRONG WIND" |
                      storms$EVTYPE == "HIGH WINDS"] <- "HIGH WIND"

storms$EVTYPE[storms$EVTYPE == "URBAN/SML STREAM FLD" |
                      storms$EVTYPE == "FLASH FLOOD"|
                      storms$EVTYPE == "FLOOD/FLASH FLOOD"] <- "FLOOD"

storms$EVTYPE[storms$EVTYPE == "WINTER WEATHER"] <- "WINTER STORM"

storms$EVTYPE[storms$EVTYPE == "WILD/FOREST FIRE"] <- "WILDFIRE"

storms$EVTYPE[storms$EVTYPE == "HEAT"] <- "EXCESSIVE HEAT"

```

## Results

### Across the United States, which types of events are most harmful with respect to population health?

The following chart shows the types of storms that cause the greatest number of injuries. Only the top ten types are shown. Tornadoes are by far the most dangerous.

```{r}

library(dplyr)
agg_injuries <- storms %>%
  group_by(EVTYPE) %>%
  summarize(count_injuries = sum(INJURIES)) %>%
  arrange(desc(count_injuries))

library(ggplot2)
ggplot(data  = head(agg_injuries, 10), aes(x = EVTYPE, y = count_injuries)) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Storm Type", y = "Number of Injuries", title = "Count of Injuries by Storm Type")

```

The following chart shows the types of storms that cause the greatest number of fatalities. Only the top ten types are shown. Tornadoes are again the most dangerous, followed by heat and flood.

```{r}

agg_fatalities <- storms %>%
  group_by(EVTYPE) %>%
  summarize(count_fatalities = sum(FATALITIES)) %>%
  arrange(desc(count_fatalities))

ggplot(data  = head(agg_fatalities, 10), aes(x = EVTYPE, y = count_fatalities)) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Storm Type", y = "Number of Fatalities", title = "Count of Fatalities by Storm Type")

```

### Across the United States, which types of events have the greatest economic consequences?

The following chart shows the types of storms that cause the greatest damages in United States dollars, including both property and crop damages. Only the top ten types are shown. Tornadoes/thunderstorm wind/hail his is by far the most costly.
```{r}

# Damages are saved as thousands (K), millions (M), and billions (B). To put these all on the same scale, multiply values.

storms$property_damage_new <- NA
storms$property_damage_new[storms$PROPDMGEXP == "K" |
                                   storms$PROPDMGEXP == "k"] <- storms$PROPDMG[storms$PROPDMGEXP == "K" |
                                   storms$PROPDMGEXP == "k"] * 1000
storms$property_damage_new[storms$PROPDMGEXP == "M" |
                                   storms$PROPDMGEXP == "m"] <- storms$PROPDMG[storms$PROPDMGEXP == "M" |
                                   storms$PROPDMGEXP == "m"] * 1000000
storms$property_damage_new[storms$PROPDMGEXP == "B" |
                                   storms$PROPDMGEXP == "b"] <- storms$PROPDMG[storms$PROPDMGEXP == "B" |
                                   storms$PROPDMGEXP == "b"] * 1000000000

storms$crop_damage_new <- NA
storms$crop_damage_new[storms$CROPDMGEXP == "K" |
                                   storms$CROPDMGEXP == "k"] <- storms$CROPDMG[storms$CROPDMGEXP == "K" |
                                   storms$CROPDMGEXP == "k"] * 1000
storms$crop_damage_new[storms$CROPDMGEXP == "M" |
                                   storms$CROPDMGEXP == "m"] <- storms$CROPDMG[storms$CROPDMGEXP == "M" |
                                   storms$CROPDMGEXP == "m"] * 1000000
storms$crop_damage_new[storms$CROPDMGEXP == "B" |
                                   storms$CROPDMGEXP == "b"] <- storms$CROPDMG[storms$CROPDMGEXP == "B" |
                                   storms$CROPDMGEXP == "b"] * 1000000000

storms$damage <- storms$property_damage_new + storms$crop_damage_new

agg_damage <- storms %>%
  group_by(EVTYPE) %>%
  summarize(total_damage = sum(damage)) %>%
  arrange(desc(total_damage))

library(scales)
ggplot(data  = head(agg_damage, 10), aes(x = EVTYPE, y = total_damage)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Storm Type", y = "Damages in US Dollars", title = "Cost of Damages by Storm Type")

```